\texttt{ensmallen}: a flexible C++ library for efficient function optimization

### 2018-10-15 doc/mlsys_paper/paper.tex
title candidate
Final first pass for me.

% TODO: title could possibly be improved
\title{\texttt{ensmallen}: a generic C++ library for fast optimization}

% Alphabetical ordering?
\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
Yannis Mentekidis
% any affiliation/email?
\And
Conrad Sanderson \\
Data61, CSIRO \\
Australia
}

@@

\begin{abstract}
\vspace*{-0.3em}
%% the abstract below still needs more meat and sharpening
We present a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and implementation of a new objective function requires at most four C++ functions. 
This can aid in the  quick implementation and prototyping of new machine learning algorithms.
In contrast to optimization approaches in existing frameworks
(such as Julia, Shogun, Vowpal Wabbit, SciPy),
the proposed Ensmallen library supports more optimizations,
more types of objective functions,
as well as seamlessly supporting user-defined objective functions and optimizers.
Empirical comparisons show that ...
TODO: briefly describe the speedups obtained in the experiments
\end{abstract}

@@

% TODO: also, we should talk up the use

@@

        % TODO: Julia
        % TODO: MATLAB paper from MLOSS 2018?

@@

In this framework, if a user wants to optimize a differentiable objective
function, they simply need to provide implementations of $f(x)$ and $f'(x)$, and
then they can use any of the gradient-based optimizers that {\tt ensmallen}
provides to perform the optimization.  Table~\ref{tab:functionality} compares
the classes of objective functions that {\tt ensmallen} can handle with other
popular frameworks and libraries.

@@

Therefore, the best we can hope to achieve is to maximize the flexibility
available, so that a user can easily implement a function $f(x)$ and have it
work with as many optimizers as possible.  For this, {\tt ensmallen} depends on
C++ policy-based design.  When implementing an objective function to be
optimized, a user can implement only a few methods and we can use C++ template
metaprogramming to check that the given functions match the requirements of the
optimizer that is being used.  When implementing an optimizer, a user can assume
that the given function to be optimized meets the required assumptions of the
optimizers, and encode those requirements as \texttt{static\_assert} checks.

For the most common case of a differentiable $f(x)$, the user must
implement only two methods:

@@

\vspace*{-1.1em}
\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|^2.
\end{equation}
\vspace*{-1.1em}

@@

Thanks to the easy abstraction, we have been able to provide support for a large
set of diverse optimizers and objective functions.  Below is a list of what is
currently available.

@@

In addition, many methods are currently in development and will be released in
the future.  % TODO: maybe some more clarity here

% not even sure this is the right term
\vspace*{-0.3em}
\section{Static polymorphism for speed}
\vspace*{-0.5em}

% TODO: better transition from the previous section

@@

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{python}
    def rosen(x): return 100.0 * (x[1] - x[2]**2.0)**2.0 + (1 - x[2])**2.0)
    res = scipy.optimize.minimize(rosen, [-1.2, 1], method='nelder-mead')
\end{minted}
\vspace*{-0.3em}

@@

Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.
% TODO: check that these make sense
% TODO: needs a reference

@@

Since all of this code generation is done at compile-time and not
at runtime, the compiler is able to perform dead code elimination % TODO: check
on the unused result, and may be able to avoid calculations entirely.  In some
cases, then, the generated {\tt Evaluate()} or {\tt Gradient()} functions may be
equivalently fast to what would be hand-written!

@@

First, we consider the added overhead of other frameworks.  For our experiment,
we choose the simple and popular Rosenbrock function~\cite{Rosenbrock1960}:
$f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2)$.  For the optimizer, we use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer.  Simulated annealing will call the objective function
very many times; for each simulation we limit the optimizer to $100,000$
objective evaluations.  We compare four frameworks for this task:

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin} from GNU Octave's {\tt optim} package
\end{itemize}
\vspace*{-0.3em}

@@

Here we can just summarize everything we've been talking about and possibly
point towards things that can happen in the future:

\begin{itemize}
  \item GPU support via Bandicoot eventually (although NVBLAS can already be
used)
  \item Additional optimizers added
  \item Anything else?
\end{itemize}

@@

The development team of {\tt ensmallen} does not include just the authors named
here but also a long list of other contributors.  See \url{%TODO about page
} for more information.

>>>>

% TODO: title could possibly be improved
%\title{\texttt{ensmallen}: a generic C++ library for fast optimization}  %% it's not clear what "optimization" refers to here
%% other possibilities:
%\title{\texttt{ensmallen}: a flexible C++ library for function optimization in machine learning}
%\title{\texttt{ensmallen}: a fast C++ library for function optimization in machine learning}
%\title{\texttt{ensmallen}: a C++ library for fast function optimization in machine learning}
%\title{\texttt{ensmallen}: a C++ library for fast and flexible function optimization}
%\title{\texttt{ensmallen}: a C++ library of fast and flexible function optimizers}
%\title{\texttt{ensmallen}: a library of flexible function optimizers in C++}
%\title{\texttt{ensmallen}: a library of fast and flexible function optimizers in C++}
%\title{\texttt{ensmallen}: a flexible C++ library for function optimization}
\title{\texttt{ensmallen}: a flexible C++ library for efficient function optimization}

% Alphabetical ordering?
% TODO: check affiliations
\author{Shikhar Bhardwaj \\
Delhi Technological University \\
Delhi, India 110042 \\
\texttt{shikhar\_bt2k15@dtu.ac.in}
\And
Ryan R. Curtin \\
RelationalAI \\
Atlanta, GA, USA 30318 \\
\texttt{ryan@ratml.org}
\And
Marcus Edel \\
Free University of Berlin \\
Arnimallee 7, 14195 Berlin \\
\texttt{marcus.edel@fu-berlin.de}
\And
Yannis Mentekidis
% any affiliation/email?
\And
Conrad Sanderson \\
Data61, CSIRO \\
Australia
}

@@

\begin{abstract}
\vspace*{-0.3em}
%% the abstract below still needs more meat and sharpening
We present {\tt ensmallen}, a fast and flexible C++ library for mathematical optimization of 
arbitrary user-supplied functions,
which can be applied to many machine learning problems.
Several types of optimizations are supported, including differentiable,
separable, constrained, and categorical objective functions.
The library provides many pre-built optimizers
(including numerous variants of SGD and Quasi-Newton optimizers)
as well as a flexible framework for implementing new optimizers and objective functions.
Implementation of a new optimizer requires only one method
and a new objective function requires at most four C++ functions. 
This can aid in the quick implementation and prototyping of new machine learning algorithms.
Due to the use of C++ template metaprogramming, {\tt ensmallen} is able to
support compiler optimizations that provide fast runtimes.
Empirical comparisons show that {\tt ensmallen} is able to outperform other
optimization frameworks (such as Julia and SciPy), sometimes by large margins.
The library is distributed under the 3-clause BSD license and is ready for use
in production environments.
\end{abstract}

@@

@@

        % Optim.jl isn't the only Julia package for optimization, but it's the
        % one we compare against.
        & Julia ({\tt Optim.jl}) \cite{julia}         & \CIRCLE & - & - & \CIRCLE & - & - & - \\

@@

In this framework, if a user wants to optimize a differentiable objective
function, they need only provide implementations of $f(x)$ and $f'(x)$, and
then they can use any of the gradient-based optimizers that {\tt ensmallen}
provides.  Table~\ref{tab:functionality} compares
the classes of objective functions that {\tt ensmallen} can handle with other
popular frameworks and libraries.

@@

Therefore, the best the library can attain is to maximize the flexibility
available, so that a user can easily implement a function $f(x)$ and have it
work with as many optimizers as possible.
To do this, {\tt ensmallen} makes heavy use of C++ template metaprogramming.
When implementing an objective function to be
optimized, a user can implement only a few methods and we can use
metaprogramming to check that the given functions match the requirements of the
optimizer that is being used.  When implementing an optimizer, we can assume
that the given function to be optimized meets the required assumptions of the
optimizers, and encode those requirements as \texttt{static\_assert} checks.

For the most common case of a differentiable $f(x)$, the user need only
implement two methods:

@@

\vspace*{-1.1em}
\begin{equation}
\operatorname{argmin}_\theta \sum_{i = 1}^n (y_i - x_i \theta)^2 =
\operatorname{argmin}_\theta \| y - X \theta \|_F^2.
\end{equation}
\vspace*{-1.1em}

@@

Due to the easy abstraction, {\tt ensmallen} is able to provide a large set of
diverse optimization algorithms.  Below is a list of what is currently
available.

@@

In addition, many optimization algorithms are currently in development and will
be released in the future.

\vspace*{-0.3em}
\section{Templates for speed}
\vspace*{-0.5em}

@@

\vspace*{-0.3em}
\begin{minted}[fontsize=\footnotesize]{python}
    def rosen(x): return 100.0 * (x[2] - x[1]**2.0)**2.0 + (1 - x[1])**2.0)
    res = scipy.optimize.minimize(rosen, [-1.2, 1], method='nelder-mead')
\end{minted}
\vspace*{-0.3em}

@@

Since {\tt ensmallen} is written in C++ with templates, the function pointer
dereferencing overhead can be avoided and the compiler can choose to inline the
function call and perform additional optimizations when possible, such as
temporary variable elimination and dead code pruning.

@@

Since all of this code generation is done at compile-time and not
at runtime, the compiler is able to perform dead code elimination
on the unused result, and may be able to avoid calculations entirely.  In some
cases, then, the generated {\tt Evaluate()} or {\tt Gradient()} functions may be
equivalently fast to what would be hand-written!

@@

First, we consider the added overhead of other frameworks.  For our experiment,
we choose the simple and popular Rosenbrock function~\cite{Rosenbrock1960}:
$f([x_1, x_2]) = 100 (x_2 - x_1^2)^2 + (1 - x_1^2)$.  For the optimizer, we use
simulated annealing~\cite{kirkpatrick1983optimization},
a gradient-free optimizer.  Simulated annealing will call the objective function
very many times; for each simulation we limit the optimizer to 100k
objective evaluations.  We compare four frameworks for this task:

@@

\vspace*{-0.3em}
\begin{itemize} \itemsep -1pt
  \item {\tt ensmallen}: only {\tt EvaluateWithGradient()} implemented
  \item {\tt ensmallen}-2: both {\tt Evaluate()} and {\tt Gradient()} separately
implemented
  \item {\tt Calculus.jl}: uses the default {\tt Calculus.jl} for
computing the gradient with {\tt Optim.jl}
  \item {\tt ForwardDiff.jl}: uses {\tt ForwardDiff.jl} for computing
the gradient with {\tt Optim.jl}
  \item {\tt Optim.jl}: implements the gradient manually
  \item {\tt scipy}: implements both objective and
gradient
  \item {\tt bfgsmin}: uses GNU Octave's {\tt optim} package
\end{itemize}
\vspace*{-0.3em}

@@

We have described {\tt ensmallen}, a flexible C++ library for function
optimization that provides an easy interface for users to implement and optimize
their desired objective functions.  Many types of functions can be optimized,
including separable and constrained functions.  The library is fast due to its
use of template metaprogramming and the Armadillo matrix library for linear
algebra.  In addition, template metaprogramming is used to make user
implementation easier by automatically generating missing methods.

GPU support for optimization is available either through the use of the NVBLAS
library~\cite{nvidia2015} or the forthcoming Bandicoot C++ GPU matrix library.
% Conrad: is it ok if we mention that here?  It's a really nice selling point
% and I am sure at least one reviewer will be thinking about GPUs.

{\tt ensmallen} is already in use as the optimization toolkit of the {\tt
mlpack} machine learning library~\cite{mlpack2018}.  For more information on the
toolkit, see the website and documentation at \url{https://www.ensmallen.org/}.

@@

The development team of {\tt ensmallen} does not include just the authors named
here but also a long list of other contributors.  See
\url{https://www.ensmallen.org/about.html} for more information.
% TODO: that URL may change
